Hydra Battles and Cie

### 2021-08-25 doc/part-hydras.tex
Alectryon:  chapter 7 (start)

\begin{Coqsrc}
Parameter Ord : Type.
Parameter lt : relation Ord.
Infix "<" := lt : schutte_scope.

Definition ordinal := Full_set Ord.
\end{Coqsrc}

@@

\begin{Coqsrc}
Variables (M:Type)
         (Lt : relation M).

Class WO : Type:=
    {
      Lt_trans : Transitive  Lt;
      Lt_irreflexive : forall a:M, ~ (Lt a a);
      well_order : forall (X:Ensemble M)(a:M),
          In X a ->
          exists a0:M, least_member  X a0
    }.
\end{Coqsrc}

\begin{Coqsrc}
  Axiom AX1 : WO lt.
\end{Coqsrc}

@@

\begin{Coqsrc}
Axiom AX2 : forall X: Ensemble Ord, 
   (exists a,  (forall y, In X y -> y < a)) ->
   countable X.

Axiom AX3 : forall X : Ensemble Ord,
              countable X -> 
              exists a,  forall y, In X y -> y < a.
\end{Coqsrc}

\texttt{AX2} and \texttt{AX3} could have been replaced by a single axiom (using the \texttt{iff} connector), but we decide to respect as most as possible the structure of Schütte's definitions.

@@

According to the  definition of a well order, every non-empty subset of \texttt{Ord} has a least element. Furthermore, this least element is unique.

\begin{Coqsrc}
Remark R : exists! z : Ord, least_member lt  ordinal z.
Proof.
  destruct inh_Ord as [a]; apply (well_order (WO:=AX1)) with a .
  split.
Qed.
\end{Coqsrc}

Assume we want to call this element  \texttt{zero}.

\begin{Coqsrc}
Definition zero : Ord.
Proof.
  Fail destruct R.
\end{Coqsrc}

\begin{Coqanswer}
The command has indeed failed with message:
Case analysis on sort Type is not allowed for inductive 
definition ex.
\end{Coqanswer}

Indeed, the basic logic of  \coq{} does not allow us to eliminate a proof of a proposition 
$\exists\,x:A,\,P(x)$ for building a term whose type lies in the sort \texttt{Type}. 
The reasons for this impossibility are explained in many documents~\cite{BC04, chlipalacpdt2011, Coq}.

@@

\begin{Coqsrc}
Class InH (A: Type) : Prop :=
   InHWit : inhabited A.

Definition some {A:Type} {H : InH A} (P: A -> Prop) := 
   epsilon (@InHWit A H) P.

Definition the {A:Type} {H : InH A} (P: A -> Prop) := 
   iota (@InHWit A H) P.
\end{Coqsrc}

In order to use these tools,  we had to tell \coq{}  that the type \texttt{Ord} is not empty:

\begin{Coqsrc}
Axiom inh_Ord : inhabited Ord.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition the_least {M: Type} {Lt}
           {inh : InH M} {WO: WO Lt} (X: Ensemble M)  : M :=
  the  (least_member  Lt X ).
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition zero: Ord :=the_least ordinal.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma zero_le (alpha : Ord) :  zero <= alpha.
Proof.
  unfold zero, the_least, the; apply iota_ind.
\end{Coqsrc}

According to the use of the description operator \texttt{iota}, we have to solve  two trivial sub-goals.
\begin{enumerate}
\item Prove that there exists a unique least member of \texttt{Ord}
\item Prove that being a least member of \texttt{Ord} entails the announced inequality 
\end{enumerate}

\begin{Coqanswer}
2 subgoals (ID 155)

  alpha : Ord
  ============================
  exists ! x : Ord, least_member lt ordinal x

subgoal 2 (ID 156) is:
 forall a : Ord, unique (least_member lt ordinal) a -> 
                a <= alpha
\end{Coqanswer}

\begin{Coqsrc}
  -  apply the_least_unicity, Inh_ord.
  -  destruct 1 as [[_ H1] _]; apply H1; split. 
Qed.
\end{Coqsrc}

@@

\begin{Coqbad}
Module Bad.

 Definition bottom := the_least (Empty_set Ord).
\end{Coqbad}

\begin{Coqanswer}
 bottom is defined
\end{Coqanswer}

@@

\begin{Coqbad}
Lemma le_zero_bottom : zero <= bottom. 
Proof. apply zero_le. Qed.

Lemma bottom_eq : bottom = bottom.
Proof. trivial. Qed.

Lemma le_bottom_zero : bottom <= zero.
Proof.
   unfold bottom, the_least, the; apply iota_ind.
\end{Coqbad}

\begin{Coqanswer}
2 subgoals (ID 413)

============================
exists ! x : Ord, least_member lt (Empty_set Ord) x

subgoal 2 (ID 414) is:
 forall a : Ord, unique (least_member lt (Empty_set Ord)) a -> 
      a <= zero
\end{Coqanswer}

\begin{Coqbad}
Abort.
End Bad.
\end{Coqbad}

@@

\begin{Coqsrc}
Definition succ (alpha : Ord) := the_least (fun beta => alpha < beta).
\end{Coqsrc}

Using \texttt{succ}, we define the folloing predicates.

\begin{Coqsrc}
Definition is_succ (alpha:Ord) := exists beta, alpha = succ beta.

Definition is_limit (alpha:Ord) := alpha <> zero /\ ~ is_succ alpha.
\end{Coqsrc}

% \begin{remark}
% Please look at remark~\vref{warning:coercions}.
% \end{remark}

@@

\begin{Coqsrc}
Definition succ_spec (alpha:Ord) :=
  least_member   lt (fun z => alpha < z).
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma succ_ok : forall alpha,  succ_spec alpha  (succ alpha).
Proof.
  intros; unfold succ, the_least, the;  apply iota_spec.
\end{Coqsrc}

\begin{Coqanswer}
1 subgoal (ID 172)

  alpha : Ord
  ============================
  exists ! x : Ord, succ_spec alpha x
\end{Coqanswer}

@@

The \coq{} proof script is quite short.

\begin{Coqsrc}
  destruct (@AX3 (Singleton _ alpha)).
  - apply countable_singleton.
  -  unfold succ_spec; apply the_least_unicity;  exists x; intuition.
Qed.     
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma lt_succ (alpha : Ord) :  alpha < succ alpha.
Proof.
  destruct  (succ_ok  alpha);  tauto.
Qed.

Hint Resolve lt_succ : schutte.

Lemma lt_succ_le (alpha beta : Ord):
  alpha < beta -> succ alpha <= beta.
Proof with eauto with schutte.
  intros  H;  pattern (succ alpha); apply the_least_ok ... 
  exists (succ alpha); red;apply lt_succ ...
Qed.
\end{Coqsrc}

\begin{Coqsrc}
Lemma lt_succ_le_2 (alpha beta : Ord):
  alpha < succ beta -> alpha <= beta.

Lemma succ_mono (alpha beta : Ord):
  alpha < beta -> succ alpha < succ beta.

Lemma succ_monoR (alpha beta : Ord) :
 succ alpha < succ beta -> alpha < beta.

Lemma lt_succ_lt (alpha beta : Ord) :
  is_limit beta ->  alpha < beta -> succ alpha < beta.
\end{Coqsrc}

>>>>

\input{movies/snippets/Schutte_basics/OrdDecl}

@@

\input{movies/snippets/Well_Orders/Mdecl}

\input{movies/snippets/Well_Orders/WODef}

\input{movies/snippets/Schutte_basics/AX1}

@@

\input{movies/snippets/Schutte_basics/AX23}

\texttt{AX2} and \texttt{AX3} could have been replaced by a single axiom (using the \texttt{iff} connector), but we decided to respect as most as possible the structure of Schütte's definitions.

@@

According to the  definition of a well order, every non-empty subset of \texttt{Ord} has a least element. Furthermore, this least element is unique. We would like to call this element  \texttt{zero}.

\input{movies/snippets/Schutte_basics/iotaDemo}

Indeed, the basic logic of  \coq{} does not allow us to eliminate a proof of a proposition 
$\exists!\,x:A,\,P(x)$ for building a term whose type lies in the sort \texttt{Type}. 
The reasons for this impossibility are explained in many documents~\cite{BC04, chlipalacpdt2011, Coq}.

@@

\input{movies/snippets/MoreEpsilonIota/Defs}

In order to use these tools,  we have to tell \coq{}  that the declared type \texttt{Ord} is not empty:

\input{movies/snippets/Schutte_basics/inhOrd}

@@

\emph{From Module\href{../theories/html/hydras.Schutte.Well_Orders.html}%
  {\texttt{Schutte.Well\_Orders}}}

\input{movies/snippets/Well_Orders/leastMemberDef}
\input{movies/snippets/Well_Orders/theLeast}

@@

\input{movies/snippets/Schutte_basics/zeroDef}

@@

\input{movies/snippets/Schutte_basics/zeroLe}

@@

\input{movies/snippets/Schutte_basics/BadBottoma}

@@

\input{movies/snippets/Schutte_basics/trivialProps}

On the other hand, the following attempt fails, because of the unprovable first sub-goal (please notice that the second sub-goal is easy to solve !).

\input{movies/snippets/Schutte_basics/Failure}

@@

\input{movies/snippets/Schutte_basics/succDef}

Using \texttt{succ}, we define the following predicates.

\input{movies/snippets/Schutte_basics/isSuccIsLimit}

% \begin{remark}
% Please look at remark~\vref{warning:coercions}.

@@

\input{movies/snippets/Schutte_basics/succSpec}

@@

\input{movies/snippets/Schutte_basics/succOka}

@@

The rest of the \coq{} proof script is quite short.

\input{movies/snippets/Schutte_basics/succOkb}

@@

\input{movies/snippets/Schutte_basics/succProps}
